THIS IS A REF OF THE TEMPLATE FILE AS WE DELETE THINGS IN THE MAIN.tex

### 2024-01-22 main_ref.tex

\twocolumn[
\icmltitle{Spurious Sentience\\
           On the Unsurprising Finding of Patterns in Latent Spaces}

@@

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, anthropomorphism, ICML}

@@

We humans are prone to seek patterns everywhere. Meaningful patterns have often proven to help us make sense of our past, navigate our present and predict the future, likely leading to evolutionary benefits even when the patterns we perceive are false \cite{foster2009evolution}. Nevertheless, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, and interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.

In statistics, misleading patterns are referred to as spurious relationships: purely associational relationships between two or more variables that are not causally related to each other. Although the definition of spuriousness varies somewhat \cite{haig2003spurious}, every student of statistics has learned to be wary that the observation of correlations do not necessarily imply causation. The world is full of these non-causal associations and as good as we as species may be at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones. Humans struggle to perceive randomness, as studies show that humans are poor at detecting e.g. \cite{falk1997making}, and creating e.g. \cite{ladouceur1996erroneous} random sequences, and even those trained in statistics and probability make inappropriate inferences, seeing patterns in data that is randomly generated \cite{zgraggen2018investigating}.

Drawing conclusions from spurious relationships is further compounded in AI research by at least two other phenomena

Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path. Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads. Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.

I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy. But I am very skeptical of claims that AI will magically provide that remedy. This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them. The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots. Since this is an opinionated post, I feel that I should start with a few disclaimers:

\begin{itemize}
    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
    \item I therefore do take issue with the way that these findings are being overblown by people with clout. Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
\end{itemize}

>>>>

\twocolumn[
\icmltitle{THIS IS A REF OF THE TEMPLATE FILE AS WE DELETE THINGS IN THE MAIN.tex}

@@

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, Anthropomorphism}

@@

We humans are prone to seek patterns everywhere. Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future. This tendency to perceive patterns likely leads to evolutionary benefits, even when the patterns we perceive are false \cite{foster2009evolution}. Nevertheless, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g. the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.

In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other. Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation. Quantitative data often show non-causal associations (see \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations}), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones. A major contributor, wherein humans struggle to conceive of actual randomness, as studies show that humans are poor at detecting e.g. \cite{falk1997making}, and creating e.g. \cite{ladouceur1996erroneous} random sequences. A common issue is a lack of expectation, that patterns tend to hint towards a causal relationship, such as correlations, may still appear at random. This leads even those trained in statistics and probability to perceive causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.

AI research and development is a perfect storm that encourages our human biases to perceive spurious sentience: patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at sentience. In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as a system that can solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience, at the level of a human or better e.g. \cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of it's release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'. When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that also aims towards human-like characteristics, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards sentience.

Present work case studies an example of claims of sentience in AI and reviews work on two human phenomena that are likely to lead to false claims of sentience: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out or biased interpretation of evidence in support of one's beliefs. We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the sentience, or lack thereof, of AI systems. We further show results from four experiments that might indicate nascent sentient 'sparks' of AGI, but that nevertheless are indications of spurious sentience.

%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path. Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads. Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.

%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy. But I am very skeptical of claims that AI will magically provide that remedy. This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them. The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots. Since this is an opinionated post, I feel that I should start with a few disclaimers:

%\begin{itemize}
%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
%    \item I therefore do take issue with the way that these findings are being overblown by people with clout. Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
%\end{itemize}

\section{Background}

\subsection{Anthropomorphism}

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}. Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. Anthropomorphism was implied because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review). Importantly, anthropomorphizing may occur independent of whether such judgments are accurate.

Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: "...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}. Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors. This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}. An illustration is the example of Tom Hank's character in the film Castaway who anthropomorphizes a ball he calls "Wilson". Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc. with human-like qualities to relieve their need for socialization, anticipating the now-forming market for apps and robots to fill this function.

The motivation as a human to be competent, so-called effectance motivation, is the most relevant to this discussion. An uncertain environment that a person cannot predict or in any way control has been shown to lead to greater anthropomorphizing. Specifically, when confronted with an incomprehensible technological agent, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience in order to faciliate learning \cite{epley2007seeing, waytz2010social}. This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn, such as the increasing demand for 'AI' skills: e.g. job postings for individuals with skills in 'GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}. Thus, individuals in advanced economies that are expected to rapidly adopt AI in various work roles may experience a pressure to rapidly learn how to effectively use and understand an opaque technology, that behaves human-like in some ways, but as it is designed can never have the cognitive functions of a human \cite{salles2020anthropomorphism, van2023reclaiming}.

\subsection{Claims of Artificial General Intelligence}

\textcolor{red}{antony: I will be adding citations and other links later, just getting text down for now.}
As previously mentioned, AGI is the current objective for which the global AI community is shooting for. As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts. This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds. True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.

Recently, Google DeepMind made claims their AlphaGeometry model reached a 'milestone' towards AGI. This model has the ability to solve complex geometry problems, which, they state it does without the need for human demonstrations during training. Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
